---
title: "Assignment 1 for Reproducible Research course"

output:
  html_document:
    keep_md: true
    
author: "Sumit Rahman"
date: "Saturday, November 14, 2015"
---


## Loading and preprocessing the data

The following code reads the data, aggregates it by day (excluding NAs) and generates a histogram showing the distribution of the daily total number of steps:

```{r, chunk1}
activity <- read.table(unz("repdata-data-activity.zip", "activity.csv"), 
                       header=T, quote="\"", sep=",")
daily.steps<-aggregate(formula=steps ~ date, FUN=sum,data=activity, na.rm=TRUE)

hist(daily.steps$steps, freq=FALSE, 
     main="total daily steps in October and November 2012", 
     xlab="daily steps")
```


## What is mean total number of steps taken per day?

Next we calculate the mean and median number of steps per day:
```{r, chunk2}
result1<-mean(daily.steps$steps)
result2<-median(daily.steps$steps)
```

The mean number of steps per day is `r format(result1,scientific=FALSE)`, and the median is `r format(result2,scientific=FALSE)`.


## What is the average daily activity pattern?
Now we take the mean number of steps for each 5 minute interval, averaged across all days, and plot this as a line graph:

```{r, chunk3}
time.steps<-aggregate(formula=steps ~ interval, FUN=mean,data=activity, 
                      na.rm=TRUE)
plot(time.steps, type="l",main="average steps per five-minute interval", 
     xlab="minutes past midnight")
```

The following code finds the interval with the largest (average) number of steps:

```{r, chunk4}
max.interval<-time.steps$interval[which(x = time.steps$steps==max(time.steps$steps),arr.ind = TRUE)]
```

The interval which has the greatest average steps taken is the five minute interval starting `r max.interval` minutes after midnight.  This average is `r round(max(time.steps$steps),digits=2)`.



## Imputing missing values

We now return to the original data set and impute the average number of steps for the corresponding 5 minute interval for any missing values.  The total number of steps for each day is then recalculated including these imputed values, and the histogram replotted:
```{r, chunk 5}
missing.steps<-is.na(activity$steps)
no.of.missing.values<-sum(missing.steps)
activity$intmean<-time.steps$steps
activity$imputed<-ifelse(missing.steps,activity$intmean,activity$steps)

daily.steps.imputed<-aggregate(formula=imputed ~ date, FUN=sum,
                               data=activity, na.rm=TRUE)

hist(daily.steps.imputed$imputed, freq=FALSE, 
     main="total daily steps in October and November 2012", 
     xlab="minutes past midnight")
```

We also recalculate the mean and median for the total daily steps.
```{r, chunk6}
result3<-mean(daily.steps.imputed$imputed)
result4<-median(daily.steps.imputed$imputed)
```
The mean number of steps per day is `r format(result3,scientific=FALSE)`, and the median is `r format(result4,scientific=FALSE)`.

The difference in the mean steps per day between the raw data and the one with imputed values is `r result3-result1`, i.e. there is no difference.  The difference in the medians is `r round(abs(result4-result2),digits=2)`, which is a very small difference.

## Are there differences in activity patterns between weekdays and weekends?

Finally, we use the imputed dataset and recalculate the average number of steps per 5 minute interval, but distinguishing between weekdays and weekends.  We then plot line graphs for both these cases.
```{r, chunk7}
require(ggplot2)
activity$day<-weekdays(as.POSIXct(activity$date))
activity$is.weekend<-ifelse(activity$day=="Saturday","weekends",
                            ifelse(activity$day=="Sunday",
                                   "weekends","weekdays"))
interval.daytype<-aggregate(activity$imputed, 
                            list(interval = activity$interval, 
                                 day = activity$is.weekend), mean)

ggplot(data=interval.daytype, aes(x=interval, y=x))+
  geom_line()+facet_grid(day~.)+
  theme(plot.title=element_text(size=rel(1.5)),
        strip.text=element_text(size=rel(2)))+
  ggtitle("average number of steps per five-minute interval \n
          for weekdays and weekends")+
  xlab("minutes past midnight")+
  ylab("average number of steps")

```


We see from this chart that during weekdays there is a much more pronounced peak in step activity corresponding to lunchtime (approximately 800 minutes past midnight).  There is also a steeper increase in weekdays (compared to weekends) at roughly 9 a.m. corresponding to the start of the working day (just after 500 minutes past midnight).
